---
title: "BIOL 4800_Fall2022_RTutorial"
output:
  word_document: default
  html_notebook: default
  pdf_document: default
---

# R Basic
  -[option] + [return] to run command that curser is in.
  -use ## to add subdividers (makes navigating easier)

## Introduction

R as calculator

```{r}
1 + 100
```

```{r}
2^2
```

```{r}
7/2
```

```{r}
4-1
```
```{r}
2*2
```

```{r}
3+5*2
```

```{r}
(3+5)*2
```

```{r}
2e2
```

```{r}
2/10000
```

```{r}
sin(1)
```

```{r}
log(1)
```

```{r}
exp(0)
```

```{r}
1==1
```

```{r}
1>=1
```

```{r}
1<1
```

```{r}
1 != 2
```

```{r}
x <- 10
x
```
x is 10
```{r}
x+x
```

```{r}
x-5
```

```{r}
x^2
```

```{r}
y <- x+x
y
```

### Vectorarization
  Vectors are sort of like lists

```{r}
1:5
```

```{r}
2^(1:5)
```
You can do transformations on whole lists of numnbers or data-sets 

```{r}
v <- 1:5
2^v
```

```{r}
log10(v)
```
You can log treansfrom whole data sets as well

### Enviorment

```{r}
ls()
```

```{r}
rm(v)
```

```{r}
ls()
```

## Package Management

```{r}
installed.packages()
```

```{r}
#install.packages("vegan", dependencies =TRUE)
```

```{r}
library(vegan)
```

```{r}
#remove.packages("vegan")
```

### Project Managment Notes

1. Set up managment structure (i.e, "data", "src", "results", "doc") -basic software everyone should have on their computer
2. Discuss avoiding redundancy of files
3. Make sure to use version control (i.e., GitHub)
## Getting Help

```{r}
?plot
```

```{r}
help(plot)
```

```{r}
?"<-"
```


```{r}
vignette("FAQ-vegan")
```

```{r}
citation("vegan")
```


## Data and their formats

```{r}
cats <- data.frame(coat = c("calico","black","tabby"),
                   weight = c(2.1,5,3.2),
                   likes_string = c(1,0,1))
```
data frames are like spread sheets

```{r}
cats
```

```{r}
getwd()
```

```{r}
write.csv(cats, file="cats.csv")
```

```{r}
cats <- read.csv("cats.csv")
```

```{r}
cats$coat
```

```{r}
cats$weight
```

```{r}
cats$weight*10
```

```{r}
log(cats$weight)
```

```{r}
logweight <- log(cats$weight)
```

```{r}
cbind(cats,logweight)
```

```{r}
cats <- cbind(cats,logweight)
```

```{r}
cats
```

```{r}
paste("My cat is ", cats$coat, ", and weighs ", cats$weight, "kg.", sep="")
```


### Data Type

```{r}
typeof(cats$coat)
```

```{r}
typeof(cats$weight)
```

```{r}
typeof(cats$likes_string)
```

```{r}
typeof(3.14159)
```

```{r}
typeof(1i)
```

```{r}
typeof(FALSE)
```

```{r}
typeof(TRUE)
```

```{r}
typeof(T)
```

```{r}
typeof(F)
```

```{r}
typeof(cats)
```

```{r}
class(cats)
```

```{r}
file.show("cats.csv")
```

```{r}
cats$likes_string <- as.logical(cats$likes_string)
typeof(cats$likes_string)
```

```{r}
cats$likes_string
```

```{r}
c("a", "b")
```

```{r}
ab <- c("a", "b")
ab
```

```{r}
c(ab,"c")
```

```{r}
c(ab, ab)
```

```{r}
1:10
```

```{r}
seq(10)
```

```{r}
z <- seq(10)
head(z, n=3)
```

```{r}
length(z)
```

```{r}
class(z)
```

```{r}
typeof(z)
```

```{r}
seq(70,100, by=5)
```


### Data Frames

```{r}
cats
```

```{r}
str(cats$coat)
```

```{r}
coats <- c("tabby", "tortoiseshell", "tortoiseshell", "black", "tabby")
coats
```

```{r}
str(coats)
```

```{r}
factor(coats)
```

```{r}
categories <- factor(coats)
class(coats)
```

```{r}
class(categories)
```

```{r}
str(categories)
```


### Lists

```{r}
list_example <- list(title="Numbers", numbers = 1:10, data=TRUE)
```

```{r}
list_example
```

```{r}
another_list <- list(1,"a",TRUE,1+1i)
another_list
```

```{r}
typeof(list_example)
```

```{r}
class(list_example)
```

```{r}
typeof(cats)
```

```{r}
class(cats)
```

```{r}
data.frame(list_example)
```

```{r}
cats
```

```{r}
cats[,3]
```

```{r}
cats[3,]
```

```{r}
cats[2:3,c(1,3)]
```


### Matricies

```{r}
matrix_example <- matrix(0, ncol=5, nrow=3)
matrix_example
```

```{r}
class(matrix_example)
```

```{r}
typeof(matrix_example)
```

```{r}
str(matrix_example)
```

```{r}
dim(matrix_example)
```

```{r}
ncol(matrix_example)
```

```{r}
nrow(matrix_example)
```

```{r}
class(data.frame(matrix_example))
```

```{r}
df_example <- data.frame(matrix_example)
df_example
```


### Subsetting

```{r}
p <- c(2.3,6.9,4.0,23,1)
p
```

```{r}
names(p) <- c('a','b','c','d','e')
p
```

```{r}
p[1]
```

```{r}
p[2:4]
```


```{r}
p[c(1,5)]
```

```{r}
p[c(1,1,1,3,5,5)]
```

```{r}
p[6]
```

```{r}
p[-3]
```

```{r}
p[c(-1,-5)]
```

```{r}
p[-(2:4)]
```

```{r}
p[c('a','c')]
```

```{r}
p[c(T,F,T,F,T)]
```

```{r}
p[names(p) != 'c']
```


### Factors

```{r}
f <- factor(c('a','b','c','d','e'))
f
```

```{r}
f[f == 'a']
```

```{r}
f[1:3]
```

```{r}
f[f %in% c('b','c')]
```

```{r}
f[-3]
```

```{r}
f2 <- factor(c('a','a','b','c','c'))
f2
```

```{r}
f2[f2 == 'a']
```

```{r}
f2[f2 %in% c('a','c')]
```


### Matricies Revisited

```{r}
set.seed(65)
m <- matrix(rnorm(6*4), ncol=4, nrow=6)
m
```

```{r}
m[3:4, c(3,1)]
```

```{r}
m[,c(3,4)]
```


### Lists Revisited

```{r}
xlist <- list(a="BIOL48006220",b=seq(1,10,by = 0.5), data="Grade")
xlist
```

```{r}
xlist[1:2]
```

```{r}
xlist['a']
```

```{r}
xlist[['a']]
```

```{r}
xlist['b']
```

```{r}
xlist[['b']]
```

```{r}
xlist$b
```


### Data Frames Revisited

```{r}
gp <- read.csv('/')
```

```{r}
head(gp, n=10L)
```

```{r}
nrow(gp)
```

```{r}
head(gp[["country"]], 10L)
```

```{r}
gp$X2001
```

```{r}
gp[c(1,3), 2:5]
```

```{r}
gp[ which(gp$X1989 <= 100 & gp$X2019 >= 100),]
```


## Conditionals and Flow

```{r}
n <- 10

if (n < 10) {
  print("n is less than 10")
} else if (n > 10) {
  print("n is greater than 10")
} else {
  print("n is equal to 10")
}
```

```{r}
for (n in seq(0,20)) {
  if (n < 10) {
    print("n is less than 10")
  } else if (n > 10) {
    print("n is greater than 10")
  } else {
    print("n is equal to 10")
  }
}
```

```{r}
g <- 0

while (g <=10) {
  print(paste(g,"is less than or equal to 10"))
  g <- g+1
}
```

## Plotting and Figures

```{r}
#install.packages("ggplot2")
```

```{r}
library(ggplot2)
```

```{r}
gp <- read.csv('gapminder_all.csv')

head(gp)
``` 

```{r}
ggplot(data=gp, mapping=aes(x=gdpPercap_1952, y=pop_1952)) + geom_point()
```

```{r}
ggplot(data=gp, mapping=aes(x=gdpPercap_2002, y=pop_2002)) + geom_point()
```

```{r}
gapminder <- read.csv("https://raw.githubusercontent.com/swcarpentry/r-novice-gapminder/gh-pages/_episodes_rmd/data/gapminder_data.csv")
head(gapminder)
```

```{r}
ggplot(data=gapminder, mapping=aes(x=gdpPercap, y=lifeExp)) + geom_point()
```

```{r}
ggplot(data=gapminder, mapping=aes(x=year, y=lifeExp, by=country, color=continent)) + geom_line()
```

```{r}
ggplot(data=gapminder, mapping=aes(x=year, y=lifeExp, by=country)) + geom_line(mapping=aes(color=continent)) + geom_point()
```

```{r}
ggplot(data=gapminder, mapping=aes(x=gdpPercap, y=lifeExp)) + geom_point(alpha=0.25) + scale_x_log10()
```

```{r}
ggplot(data=gapminder, mapping=aes(x=gdpPercap, y=lifeExp)) + geom_point(alpha=0.25, color="purple", size = 0.9) + scale_x_log10() + geom_smooth(method = lm, color="gold", size = 1.25)
```

```{r}
africas <- gapminder[gapminder$continent == "Africa", ]
head(africas)
```

```{r}
ggplot(data=africas, mapping = aes(x=year, y=lifeExp)) +
  geom_line(color = "red") +
  facet_wrap( ~ country) +
  theme(axis.text.x = element_text(angle = 90)) +
  labs(
    x = "Year",
    y = "Life Expectency",
    title = "Life Expectency Over Time in African Countries"
  )
```

```{r}
AfricanLifeExp <- ggplot(data=africas, mapping = aes(x=year, y=lifeExp)) +
  geom_line(color = "red") +
  facet_wrap( ~ country) +
  theme(axis.text.x = element_text(angle = 90)) +
  labs(
    x = "Year",
    y = "Life Expectency",
    title = "Life Expectency Over Time in African Countries"
  )
```

```{r}
ggsave(filename = "AfricanLifeExp.png", plot = AfricanLifeExp, width = 24, height = 40, dpi = 300, units = "cm")
```

```{r}
pdf(file = "AfricanLifeExp.pdf", width = 24, height = 40)
plot(AfricanLifeExp)
dev.off()
```

```{r}
write.table(gapminder, file="gapminder_web.csv", sep=",")
```

```{r}
write.csv(africas, file="gapminder_web_africas.csv")
```


### Fancy Plots

```{r}
install.packages(c("ggridges","viridis","hrbrthemes"), dependencies = T)
```

```{r}
library(ggridges)
library(ggplot2)
library(viridis)
```

```{r}
library(hrbrthemes)
```

```{r}
#Plot
ggplot(lincoln_weather, aes(x = `Mean Temperature [F]`, y = `Month`, fill = ..x..)) +
  geom_density_ridges_gradient(scale = 3, rel_min_height = 0.01) +
  scale_fill_viridis(name = "Temp. [F]", option = "C") +
  labs(title = 'Temperatures in Lincoln NE in 2016') +
  theme_ipsum() +
    theme(
      legend.position="none",
      panel.spacing = unit(0.1, "lines"),
      strip.text.x = element_text(size = 8)
    )
```

```{r}
#spider_data <- read.csv(file="https://wsc.nmbe.ch/resources/species_export_20221101.csv", header=TRUE)
#write.csv(spider_data, "data/spider_data_20221101.csv")

spider_data <- read.csv("data/spider_data_20221101.csv")
spider_data
```

```{r}
library(ggridges)
library(ggplot2)
library(viridis)
library(hrbrthemes)

# Plot
ggplot(spider_data[1:5000,], aes(x = `year`, y = `family`)) +
  geom_density_ridges(alpha=0.6, stat="binline", bins=20) +
  #geom_density_ridges_gradient(scale = 3, rel_min_height = 0.01) +
  labs(title = 'Spider species descriptions over time by family') +
  xlab(label = "Species Description Year") +
  ylab(label = "Spider Family") +
  theme_ridges() +
    theme(
      legend.position="none",
      panel.spacing = unit(0.1, "lines"),
      strip.text.x = element_text(size = 8)
    )
```

```{r}
#install.packages(c("ggstatsplot","palmerpenguins","tidyverse"), dependencies=T)
library(ggstatsplot)
```

```{r}
library(tidyverse)
```

```{r}
data("penguins", package = "palmerpenguins")
penguins
```

```{r}
penguins2 <- drop_na(penguins)
penguins2
```

```{r}
plt <- ggbetweenstats(
  data = penguins,
  x = species,
  y = bill_length_mm
)
```

```{r}
plt
```

```{r}
gapminder2 <- drop_na(gapminder)
```

```{r}
boxplot(gapminder2$lifeExp ~ gapminder2$continent)
```

```{r}
#install.packages("vioplot", dependencies = T)

library(vioplot)
```

```{r}
with(gapminder2, vioplot(
  lifeExp~continent, col = "blue"
))
```

```{r}
plt <- ggbetweenstats(
  data = gapminder2,
  x = continent,
  y = lifeExp
)
```

```{r}
plt
```

```{r}
plt2 <- plt  +
  theme(
    axis.ticks = element_blank(),
    axis.line = element_line(colour = "grey50"),
    panel.grid = element_line(color = "#b4aea9"),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_line(linetype = "dashed"),
    panel.background = element_rect(fill = "#fbf9f4", color = "#fbf9f4"),
    plot.background = element_rect(fill = "#fbf9f4", color = "#fbf9f4")
  ) +
  labs(
    x = "Continent",
    y = "Life Expectency (years)",
    title = "Life expectency of people living on each continent"
  )
```

```{r}
plt2
```

```{r}
ggsave(
  filename = "results/gapminder_lifeExpXcontinentweb-violinplot-with-ggstatsplot.png",
  plot = plt2,
  width = 8,
  height = 8,
  device = "png"
)
```

```{r}
# Get the shape file of Africa
#install.packages("maptools")
library(maptools)
```

```{r}
data(wrld_simpl)
afr<-wrld_simpl[wrld_simpl$REGION==2,]
 
# We can visualize the region's boundaries with the plot function
plot(afr)
```

```{r}
levels(penguins2$species)
```

```{r}
penguin_matrix <- with(penguins2, cbind(bill_length_mm, bill_depth_mm, flipper_length_mm, body_mass_g))
penguin_matrix
```

```{r}
penguin_pca <- princomp(penguin_matrix, cor=TRUE)
summary(penguin_pca)
```

```{r}
loadings(penguin_pca)
```

```{r}
biplot(penguin_pca, xlab=penguins2[,2])
```

```{r}
penguin_pca$scores
```


## Statistics

### Additional Ways of Importing/Reading Data

```{r}
rand <- c(12, 54, 98, 65, 38)
rand
```

```{r}
sum(rand)
```

```{r}
length(rand)
```

```{r}
avg.rand <- sum(rand)/length(rand)
avg.rand
```

```{r}
mean(rand)
```

```{r}
sort(rand)
```

```{r}
min(rand)
```

```{r}
max(rand)
```

```{r}
cumsum(rand)
```

```{r}
diff(rand)
```

```{r}
rand[2]
```

```{r}
rand*2
```

```{r}
pedes <- scan()
```

```{r}
pedes
```



### Analyses

#### Univariate Statistics

1. Catagorial Data
  a. Barplots
  
```{r}
beer <- c(3, 4, 1, 1, 3, 4, 3, 3, 1, 3, 2, 1, 2, 1, 2, 3, 2, 3, 1, 1, 1, 1, 4, 3, 1)
length(beer)
```
  
```{r}
barplot(beer) # WRONG
```
  
```{r}
table(beer)
```

```{r}
barplot(table(beer), xlab="Beer", ylab="Frequency") # Counts
```

```{r}
barplot(table(beer)/length(beer), xlab="Beer", ylab="Proportion") # Proportions instead of frequencies
```

  b. Pie Charts
  
```{r}
pie(table(beer), main="Beer preference by students")
```

2. Numerical Data
  a. Stem-and-Leaf Plots
  
```{r}
#stem(pedes)
```
  
  b. Strip Chart
  
```{r}
#stripchart(pedes, method="stack")
```
  
3.Measures of Center
  a. Mean
  
```{r}
mean(pedes)
```
  
```{r}
mean(gapminder2$lifeExp)
```

  b.Median
  
```{r}
median(pedes)
```

```{r}
median(gapminder$lifeExp)
```

```{r}
with(gapminder, median(lifeExp))
```

  c.Mode
```{r}
which(table(pedes) == max(table(pedes)))
```

4. Variation
  a. Range
  
```{r}
range(pedes)
```

```{r}
diff(range(pedes))
```

  b. Variance
  
```{r}
var(pedes)
```

```{r}
sd(pedes)
```

  c. IQRs (Interquartile Ranges)
  
```{r}
IQR(pedes)
```
  
  d. Z-Scores
  
```{r}
scale(pedes)
```

  e. Summary
  
```{r}
summary(pedes)
```
  
5. Plot
  a. Histograms
  
```{r}
#hist(pedes)
```

```{r}
#hist(pedes, breaks="scott")
```

```{r}
#hist(pedes, breaks="scott", prob=TRUE)
```

```{r}
#hist(pedes, breaks="scott", prob=T)
#lines(density(pedes))
```

```{r}
#plot(density(pedes))
```

  b. Box Plots
```{r}
#boxplot(pedes)
```

```{r}
summary(pedes)
```

#### Bivariate and Multivariate Statistics

1. Plotting amnd Regrassion
  a. Box Plotting
  
```{r}
spid.gen <- read.csv("spider_genitalia.csv", header = "T")
spid.gen
```

```{r}
boxplot(spid.gen$left.bulb/spid.gen$carapace.length ~ spid.gen$habitat)
```

  b. Scatter Plotting
  
```{r}
plot(spid.gen$left.bulb ~ spid.gen$right.bulb, pch = as.character(spid.gen$habitat))
```

  c. Linear Regression
  
```{r}
gen.reg <- lm(spid.gen$left.bulb ~ spid.gen$right.bulb)
gen.reg
```
  
```{r}
summary(gen.reg)
```

```{r}
plot(spid.gen$left.bulb ~ spid.gen$right.bulb)
abline(gen.reg)
```

  d. Correlation Coefficients & Spearman Rank Correlation
  
```{r}
cor.gen <- with(spid.gen, cor(left.bulb, right.bulb))
cor.gen
```

```{r}
cor.gen^2
```
this creates the R value

```{r}
spearman.cor.gen <- with(spid.gen, cor(left.bulb, right.bulb, method = "spearman"))
spearman.cor.gen
```

  e. Residuals
  
```{r}
residuals(gen.reg)
```
  
 f. Transformations
 
```{r}
plot(spid.gen$left.bulb~spid.gen$right.bulb)
```
 
2. Comparing Discrete Treatment Effects

  a. Chi-squared
  
```{r}
obs_weighted <- c(4,15,6,15,18,2) 
obs_fair <- c(10,10,10,10,10,10)
exp <- c(.16,.17,.16,.17,.17,.17)
chisq.test(obs_weighted, p=exp)
chisq.test(obs_fair, p=exp)
```
  
  b. T-test
  
```{r}
#H0 - true mean is equal to 0, but you can set true mean to other values with "mu="
t.test(spid.gen$carapace.length)

t.test(spid.gen$carapace.length, mu=29)
```
  
```{r}
habitat.t2 <- t.test(spid.gen$carapace.length~spid.gen$habitat)
habitat.t2
```
  This asks if the spiders carapaces (mid section of body size) are diffdrent sizews based on habit/location
  
```{r}
habitat.tless <- t.test(spid.gen$left.bulb, alternative = "less")
habitat.tless
```
 
```{r}
habitat.tgreater <- t.test(spid.gen$left.bulb, alternative = "greater")
habitat.tgreater
```
  
  c. Analysis of Variance (ANOVA)
  
```{r}
gen.lm <- lm(spid.gen$left.bulb~spid.gen$habitat)
```
  
```{r}
gapminder.turkey.two.way <- TukeyHSD(gapminder.lifeexp.continent.anova)
gapminder.turkey.two.way
```
  
```{r}
plot(1:25, rep(0.25,25), pch=1:25, col=1:25, ylim = c(0,6), cex=2, ylab = "Line types(lty) 1 to 6", xlab = "Plotting character (pch) 1 to 25 and colours (col) 1 to 8", main = "Line types (lty), plotting characters (pch),\nand colors (col) for plot and xyplot", lab=c(27,7,2))

points(1:8, rep(0.5,8), pch=20, col=1:8, cex=3)

abline(h=1:6, lty=1:6, col=1:6, lwd=3)
```
  
```{r}
pdf("./results/PlotCheatSheet.pdf")

plot(1:25, rep(0.25,25), pch=1:25, col=1:25, ylim = c(0,6), cex=2, ylab = "Line types(lty) 1 to 6", xlab = "Plotting character (pch) 1 to 25 and colours (col) 1 to 8", main = "Line types (lty), plotting characters (pch),\nand colors (col) for plot and xyplot", lab=c(27,7,2))

points(1:8, rep(0.5,8), pch=20, col=1:8, cex=3)

abline(h=1:6, lty=1:6, col=1:6, lwd=3)

dev.off()
```
  
## Looping and Conditionals

a. If then statements
```{r}
x <- 0
if (x > 0){
  print("Positive Number")
} else if (x < 0) {
  print("Negative Number")
} else
  print("Zero")
```

```{r}
a <- c(5,7,2,9)
ifelse(a %% 2 == 0, "even", "odd")
```
  
b. For Loops
```{r}
z <- c(2,5,3,9,8,11,6)
count <- 0

for (value in z){
  if(value %% 2 == 0) count = count+1
}
print(count)
```

c. While Statement
```{r}
i <- 1
while (i < 5){
  print(1)
  i = i+1
}
```

d. Interruptions

```{r}
x <- 1:5

for (value in x) {
  print(value)
  if (value ==3) {
    #print(value)
    break
  }
  #print(value)
}
```
  
```{r}
x <- 1:5
for(value in x) {
  if (value == 3) {
    next
  }
  print(value)
}
```
  Next skios the value 3, because you told i too in the if statement
  
```{r}
x <- 1
repeat {
  print(x)
  x = x+1
  if (x ==21) {
    break
  }
}
```
  
## Randomization and Dataset Management (short)
a. Dataset Managment
```{r}
B <- matrix(
  c(2,4,3,1,5,87),
  nrow=3,
  ncol=2
)

B
```

```{r}
t(B)
```
  
```{r}
C <- matrix(
  c(7,4,2),
  nrow=3,
  ncol=1
)

C
```
  
```{r}
BC <- cbind(B,C)

BC
```

```{r}
c(B)
```
  
b. Randomizations

```{r}
# rnorm(n,mean,sd)
# sample(x, size, replace = FALSE, prob = NULL)
```

```{r}
ndist <- rnorm(100, 50, 10)
print(ndist)
```

```{r}
mean(ndist)
sd(ndist)
```

```{r}
hist(ndist)
```
  
```{r}
sample(ndist, 10)
```
  sample(whatyoursampleing, n random of how much you want to sample)
  
```{r}
test <- 1:100
test
```
  
```{r}
 sample(test, 50, replace= FALSE)
```
  
```{r}
sample(test, 50, replace=TRUE)
```
  
  

  
  
  
  
  
  
  
  
  
  
  
  



